---
title: "Course Project 1 Fitness Tracker"
author: "Michaela Spiegel"
date: "11 April 2018"
output: md_document
---

## Loading and preprocessing the data 

First the data of a fitness tracker is loaded.

It contains three columns, the number of steps, the date and interval.

The data was recorded in October and November 2012 
in 5 minute intervals, which leads to 288 intervals per day

```{r load data, echo = TRUE}
activity <- read.csv("activity.csv")
head(activity,3)
str(activity)
```
The date column is transformed into a Date format.

```{r preprocess data, echo = TRUE}
library(lubridate)
activity$date <- ymd(activity$date)
activity$day <- weekdays(activity$date)
activity$month <- month(activity$date)
activity$year <- year(activity$date)
str(activity)
```
## What is the mean total number of steps taken per day?

The steps taken on every day, were measured in 5 minute intervals.
Now they are summed up for each day.


```{r mean steps, echo = TRUE}
steps_per_day <- aggregate(activity$steps, by = list(activity$date), FUN = sum, na.rm = TRUE)
names(steps_per_day) <- c("date", "steps")
head(steps_per_day, 3)
```
The histgram shows the frequencies of total steps per day 
in intervals of 5,000 steps. 

```{r hist mean steps, echo = TRUE}
png(filename = "figure/hist_total_steps_per_day_without_imputation.png")
hist(steps_per_day$steps,col = "indianred", main = "histogram of total steps per day", xlab = "Steps per Day")
dev.off()
hist(steps_per_day$steps,col = "indianred", main = "histogram of total steps per day", xlab = "Steps per Day")
```

Most often between 10,000 and 15,000 steps
were taken.

The summary shows us that the mean number of total steps per
day is 9,354 steps and the median 10,395 steps.
```{r summary mean steps, echo = TRUE}
summary(steps_per_day$steps)
```

## What is the average daily activity pattern?

```{r time series}
mean_steps_interval <- aggregate(activity$steps, by = list(activity$interval), FUN = mean, na.rm=TRUE)
names(mean_steps_interval) <- c("interval", "mean_steps")
png(filename = "figure/mean_steps_per_day_without_imputation.png")
plot(mean_steps_interval, type="l")
dev.off()
plot(mean_steps_interval, type="l")
```

This figure shows the average daily activity pattern, which is measured in 288 interval steps each for 5 minutes. The x-axis shows the time, where 500 
corresponds to 5 am and 20000 to 8 pm.

Now I'm aranging the data by the number of steps.

```{r}
head(mean_steps_interval, 3)
```


```{r max daily activity}
library(dplyr)
mean_steps_interval <- tbl_df(mean_steps_interval)
mean_steps_interval<- mean_steps_interval %>% arrange(desc(mean_steps))
head(mean_steps_interval,5)
```
The highest number of average steps with 206 is in the time interval
between 8:35 and 8:40 am. 

The five highest average number of steps are in the
25 minute interval between 8:30-8:55 am.


## Imputing missing values

The data contains a lot of missing values (NA's), in total 2304, which is 13 %

```{r missing values}
nr_missing <- 0 
nr_missing <- sum(is.na(activity$steps))
nr_missing*100/nrow(activity)
```


I first get the indices of the missing values and then loop over 
them and impute mean values. I save everything in a new datframe for comparison later.

```{r filling missing values}
activity_impute <- data.frame(activity)
index_na <- which(is.na(activity$steps))
for(i in index_na){
  activity_impute$steps[i] <- mean_steps_interval$mean_steps[mean_steps_interval$interval == activity$interval[i]]
  #print( mean_steps_interval$mean_steps[mean_steps_interval$interval == activity$interval[i]])
}
```

Test if it worked..


```{r}
sum(is.na(activity_impute$steps))
```
So all the missing values are now replaced by the mean values of their time interval.

## Identifiying the impact of imputation

In the next step the histogram of total number of steps per day is calculated again, as before.
```{r}
steps_per_day_impute <- aggregate(activity_impute$steps, by = list(activity_impute$date), FUN = sum, na.rm = TRUE)
names(steps_per_day_impute) <- c("date", "steps")
head(steps_per_day_impute, 3)
png(filename = "figure/compare_imputation_hist_total_steps_per_day.png")
par(mfrow=c(1,2))
hist(steps_per_day_impute$steps,col = "indianred", main = "after imputation: total steps / day", xlab = "Steps per Day")
hist(steps_per_day$steps,col = "indianred", main = "before: total steps / day", xlab = "Steps per Day")
dev.off()
par(mfrow=c(1,2))
hist(steps_per_day_impute$steps,col = "indianred", main = "after imputation: total steps / day", xlab = "Steps per Day")
hist(steps_per_day$steps,col = "indianred", main = "before: total steps / day", xlab = "Steps per Day")
```

The histogram looks quite different, than before... The peak is still between 10,000 and 15,000.
But the missing values in the number of steps leads to a lot days where under 10,000 steps were taken, just because the values are missing.
So it is the best idea to impute mean values for these times in order to get a reasonable analysis.

The mean number of steps before imputation was 9,354 steps and is now 10,766 steps. The median number of steps was without imputation 10,395 steps and extended now to 10,766 steps.

```{r}
summary(steps_per_day_impute$steps)
```

## Are there differences in activity patterns between weekdays and weekends?

To answer this question we devide the dataset into weekday and weekend data first.

```{r}
weekend = c("Samstag", "Sonntag")
activity_weekend <- (activity_impute[activity_impute$day %in% weekend,])
activity_weekdays <- (activity_impute[!activity_impute$day %in% weekend,])
```

Afterwards we calculate the time series plot, 
of average number of steps in time intervals of 5 minutes again to
compare wether the activity on weekends differ from the mean activity during the week.
```{r}
mean_steps_interval_weekend <- aggregate(activity_weekend$steps, by = list(activity_weekend$interval), FUN = mean)
mean_steps_interval_weekdays <- aggregate(activity_weekdays$steps, by = list(activity_weekdays$interval), FUN = mean)
names(mean_steps_interval_weekend) <- c("interval", "mean_steps")
names(mean_steps_interval_weekdays) <- c("interval", "mean_steps")
png(filename = "figure/mean_steps_per_day_weekends_vs_weekdays.png")
par(mfrow=c(1,2))
plot(mean_steps_interval_weekdays, type="l")
plot(mean_steps_interval_weekend, type="l")
dev.off()
par(mfrow=c(1,2))
plot(mean_steps_interval_weekdays, type="l")
plot(mean_steps_interval_weekend, type="l")
```

We can see much more activity during the day for weekends and also some activity late in the night.


