---
title: "Course Project 1 Fitness Tracker"
author: "Michaela Spiegel"
date: "5 Januar 2018"
output: html_document
---
## Loading and preprocessing the data 

First the data of a fitness tracker is loaded.

It contains three columns, the number of steps, the date and interval.

The data was recorded in October and November 2012 
in 5 minute intervals, which leads to 288 intervals per day

```{r load data, echo = TRUE}
activity <- read.csv("activity.csv")
head(activity,3)
str(activity)
```
The date column is transformed into a Date format.

```{r preprocess data, echo = TRUE}
library(lubridate)
activity$date <- ymd(activity$date)
#activity$year <- year(activity$date)
str(activity)
```
## What is the mean total number of steps taken per day?

The steps taken on every day, were measured in 5 minute intervals.
Now they are summed up for each day.


```{r mean steps, echo = TRUE}
steps_per_day <- aggregate(activity$steps, by = list(activity$date), FUN = sum, na.rm = TRUE)
names(steps_per_day) <- c("date", "steps")
```
The histgram shows the frequencies of total steps per day 
in intervals of 5,000 steps. 

```{r hist mean steps, echo = TRUE}
hist(steps_per_day$steps,col = "indianred", main = "histogram of total steps per day", xlab = "Steps per Day")
```

Most often between 10,000 and 15,000 steps
were taken.

The summary shows us that the mean number of total steps per
day is 9,354 steps and the median 10,395 steps.
```{r summary mean steps, echo = TRUE}
summary(steps_per_day$steps)
```

## What is the average daily activity pattern?

```{r time series}
mean_steps_interval <- aggregate(activity$steps, by = list(activity$interval), FUN = mean, na.rm=TRUE)
names(mean_steps_interval) <- c("interval", "mean steps")
plot(mean_steps_interval, type="l")
```


## Imputing missing values

## Are there differences in activity patterns between weekdays and weekends?